\section{Integration Model and Architecture}\label{sec:arch}

\subsection{The Integration Model}

In xlog a neural network \emph{is} a predicate. A declaration binds a network to a predicate symbol; evaluating that predicate runs a forward pass, and the network's softmax over a label set becomes a distribution over weighted facts. Those facts enter whichever reasoning mode the program uses---deterministic joins, probabilistic knowledge compilation, or epistemic world-view search---through the same relational interface, and the gradient of a query's value with respect to the network outputs flows back across the same boundary. The integration is \emph{uniform} (one mechanism, independent of backend) and \emph{transfer-free} (the facts, the compiled circuit, and the gradients never leave the GPU). The rest of this section describes the substrate that makes both properties hold: a strict GPU-residency contract, a layered crate architecture, a compilation pipeline, and an extensible IR stack.

\subsection{GPU Residency Model}

xlog enforces a hard guarantee: all runtime semantic state resides in GPU device memory, or the system returns a deterministic error. There is no silent out-of-core spilling and no implicit CPU fallback; a workload that exceeds the memory budget raises \texttt{RESOURCE\_EXHAUSTED} with diagnostics rather than degrading transparently. The contract extends to production query paths, which target zero device-to-host (D2H) transfers: the kernel provider accounts for transfers at byte granularity through atomic counters, so a performance-critical section can be bracketed and asserted to download nothing---the check the training path relies on to prove its gradient path stays on-device.

The payoff is bandwidth asymmetry. PCIe~4.0 $\times16$ delivers roughly 32\,GB/s, while GPU HBM provides 1--3\,TB/s. A single unnecessary D2H round-trip for a moderate relation (10M tuples at 16 bytes, ${\approx}160$\,MB) costs about 5\,ms---enough to dominate a semi-naive iteration that completes in microseconds on-device. Keeping fact stores, deltas, circuit values, solver state, and gradient buffers exclusively on the GPU eliminates this class of bottleneck; host involvement is restricted to orchestration, I/O, and compilation.

\subsection{Crate Architecture}

\begin{figure*}[tbp]
  \centering
  \begin{tikzpicture}[x=1mm,y=1mm,
      cell/.style={fnode,anchor=north west,font=\footnotesize,
                   inner xsep=2mm,inner ysep=1.6mm,minimum height=12mm},
      tag/.style={ftier,anchor=north west,text width=21mm,minimum height=12mm,
                  font=\footnotesize}]
    % ---- Tier 4 --- interfaces (host-facing) ----
    \node[tag] (t4tag) at (0,0)
      {\textbf{Tier 4}\\{\scriptsize\scshape interfaces}};
    \node[cell,text width=67.5mm] (t4c1) at (27,0)
      {Python API (PyO3)\\[0.6mm]\cratechip{pyxlog}};
    \node[cell,text width=67.5mm] (t4c2) at (100.5,0)
      {command-line interface\\[0.6mm]\cratechip{xlog-cli}};
    % ---- Tier 3 --- pipelines ----
    \node[tag] (t3tag) at (0,-19.5)
      {\textbf{Tier 3}\\{\scriptsize\scshape pipelines}};
    \node[cell,text width=43mm] (t3c1) at (27,-19.5)
      {deterministic\\[0.6mm]\cratechip{xlog-gpu}};
    \node[cell,text width=43mm] (t3c2) at (76,-19.5)
      {probabilistic $\cdot$ verified KC\\[0.6mm]\cratechip{xlog-prob}};
    \node[cell,text width=43mm] (t3c3) at (125,-19.5)
      {neural-symbolic\\[0.6mm]\cratechip{xlog-neural} \cratechip{xlog-induce}};
    % ---- Tier 2 --- subsystems ----
    \node[tag] (t2tag) at (0,-36.5)
      {\textbf{Tier 2}\\{\scriptsize\scshape subsystems}};
    \node[cell,text width=43mm] (t2c1) at (27,-36.5)
      {language frontend\\[0.6mm]\cratechip{xlog-logic}};
    \node[cell,text width=43mm] (t2c2) at (76,-36.5)
      {GPU runtime executor\\[0.6mm]\cratechip{xlog-runtime}};
    \node[cell,text width=43mm] (t2c3) at (125,-36.5)
      {GPU CDCL $\cdot$ local search\\[0.6mm]\cratechip{xlog-solve}};
    % ---- Tier 1 --- IR & device services ----
    \node[tag] (t1tag) at (0,-53.5)
      {\textbf{Tier 1}\\{\scriptsize\scshape ir \& device\\[-0.4mm]\scriptsize\scshape services}};
    \node[cell,text width=43mm] (t1c1) at (27,-53.5)
      {IR stack: RIR/PIR/XGCF/EIR\\[0.6mm]\cratechip{xlog-ir}};
    \node[cell,text width=43mm] (t1c2) at (76,-53.5)
      {CUDA wrapper, Arrow/DLPack\\[0.6mm]\cratechip{xlog-cuda}};
    \node[cell,text width=43mm] (t1c3) at (125,-53.5)
      {runtime statistics\\[0.6mm]\cratechip{xlog-stats}};
    % ---- Tier 0 --- core ----
    \node[tag] (t0tag) at (0,-70.5)
      {\textbf{Tier 0}\\{\scriptsize\scshape core}};
    \node[cell,text width=141mm] (t0c1) at (27,-70.5)
      {scalar types $\cdot$ kernel-provider trait $\cdot$ symbol table $\cdot$ memory budgets\\[0.6mm]\cratechip{xlog-core}};
    % ---- GPU device plane behind tiers 1-3 ----
    \begin{scope}[on background layer]
      \node[fplane,fit=(t3tag)(t3c3)(t1tag)(t1c3),inner sep=2.2mm] (plane) {};
    \end{scope}
    \node[fplanetitle,anchor=east] at ([xshift=-4mm]plane.north east)
      {\planetitle{runtime semantic state $\cdot$ gpu-resident}};
    % ---- dependency arrows (tier-level, down the tag column) ----
    \draw[farrow] (12.5,-12) -- (12.5,-19.5);
    \draw[farrow] (12.5,-31.5) -- (12.5,-36.5);
    \draw[farrow] (12.5,-48.5) -- (12.5,-53.5);
    \draw[farrow] (12.5,-65.5) -- (12.5,-70.5);
    \node[flabel,anchor=west] at (14,-14.2) {depends on};
    % ---- host annotations ----
    \node[flabel,anchor=south east] at (172,1.5)
      {host: orchestration $\cdot$ I/O $\cdot$ compilation};
    \node[flabel,anchor=north east] at (172,-84)
      {host-side foundation: shared types \& traits};
  \end{tikzpicture}
  \caption{xlog crate hierarchy, with the workspace's production crates in
  place. Each tier depends only on lower tiers; the shaded device plane marks
  the tiers whose runtime semantic state is GPU-resident, while the host
  provides orchestration, I/O, and compilation. Test-only crates
  (\texttt{xlog-cuda-tests}, \texttt{xlog-integration}) are omitted.}
  \label{fig:arch}
\end{figure*}

The workspace is organized into five tiers with strict layering: no crate depends on a peer or a higher tier (\Cref{fig:arch}). A leaf \emph{core} crate defines the shared scalar types, the kernel-provider trait, error and memory-budget types, and a bidirectional symbol table for dictionary-encoded strings. Above it, a tier of intermediate representations and device services wraps CUDA via cudarc~\cite{cudarc}, stages kernel modules (cubin-first with portable PTX fallback), and exposes Arrow and DLPack interoperability. The three core subsystems---the typed language frontend (\Cref{sec:lang}), the GPU runtime executor, and the GPU CDCL/local-search solver---compose into integrated pipelines for deterministic, probabilistic, and neural-symbolic execution, which a PyO3~\cite{pyo3} extension module (\texttt{pyxlog}) and a command-line binary surface to users.

\subsection{Compilation Pipeline}

A program is compiled to a GPU-executable plan in five stages. \textbf{Parsing} (a PEG parser built with pest) produces an AST covering the full surface, including probabilistic facts, annotated disjunctions, and neural-predicate declarations. \textbf{Stratification} builds a predicate dependency graph with positive, negative, and aggregate edges and computes its strongly connected components (SCCs) with Tarjan's algorithm; any SCC that crosses a negative or aggregate edge is rejected, guaranteeing a unique fixpoint and a deterministic stratum order. \textbf{Lowering} translates each rule into a relational-IR tree---body literals become scans, shared variables become join keys, negated literals become anti-joins, and recursive SCCs are wrapped in fixpoint nodes. \textbf{Optimization} applies cost-based predicate pushdown and join ordering, promotes eligible multiway rules to the WCOJ subsystem, and rewrites bound recursive queries via magic sets (\Cref{sec:datalog}). \textbf{Plan assembly} emits an execution plan holding SCCs in dependency order, which the runtime interprets by dispatching each node to its CUDA kernel.

\subsection{IR Stack}

A layered IR stack keeps the backends decoupled; each level has one role.

\textbf{RIR (Relational IR)} is the algebraic tree---\texttt{Scan}, \texttt{Filter}, \texttt{Project}, \texttt{Join}, \texttt{GroupBy}, \texttt{Union}, \texttt{Distinct}, \texttt{Diff}, \texttt{Fixpoint}, and the worst-case-optimal \texttt{MultiWayJoin}/\texttt{ChainJoin}---carrying cardinality, skew, and delta-vs-full metadata. Every backend consumes it.

\textbf{PIR (Provenance IR)} is the weighted Boolean formula (\texttt{Lit}, \texttt{NegLit}, \texttt{And}, \texttt{Or}, \texttt{Decision}) derived from provenance tracking over a probabilistic program, capturing its structure without execution order.

\textbf{XGCF (xlog GPU Circuit Format)} is the compiled, device-resident form: a levelized DAG whose level offsets let every node at one topological level evaluate in a single kernel launch, with forward (log-space weighted model counting) and backward (adjoint) passes.

\textbf{EIR (Epistemic IR)} preserves modal operators for world-view reasoning and lowers to a dedicated GPU execution plan rather than ordinary relational algebra (\Cref{sec:epistemic}).
