\section{The xlog Language}\label{sec:lang}

This section describes the xlog surface: types, expressions, user-defined functions, modules, and stratified aggregation. Probabilistic facts (\texttt{p::f.}) and neural predicate declarations (\texttt{nn/k}) are language-level constructs covered in \Cref{sec:prob,sec:nesy}.

\subsection{Design Principles}

Three principles shape the surface, each a deliberate enabler of GPU execution rather than a concession. \textbf{Typed predicates:} every predicate has a declared signature over a closed set of scalar types (\texttt{u32}, \texttt{u64}, \texttt{i32}, \texttt{i64}, \texttt{f32}, \texttt{f64}, \texttt{bool}, \texttt{symbol}), so argument mismatches are caught before any kernel is generated and allocation stays bounded and columnar. \textbf{Stratified semantics:} negation, aggregation, and recursion interact through an SCC analysis on the predicate dependency graph, and programs whose SCCs cross negation or aggregation are rejected at compile time, guaranteeing a unique fixpoint computable by a deterministic semi-naive schedule. \textbf{One language for many paradigms:} probabilistic facts, annotated disjunctions, neural predicates, and SAT constraints share the syntactic core of deterministic Datalog; parsing, dependency analysis, and most lowering are shared, while backend-specific passes refine the plan. A language admitting unbounded terms or recursion through negation would forfeit exactly the allocation and fixpoint guarantees the GPU runtime depends on.

A minimal program exhibits the core surface:

\lstinputlisting[style=xlogstyle,
                 caption={Minimal xlog program: typed predicates, facts, a recursive rule, a query.},
                 label={lst:hello}]{examples/01_hello.xlog}

\subsection{Type System}\label{sec:lang:types}

Every predicate is declared with \texttt{pred} over the eight scalar types, with \texttt{f32}/\texttt{f64} following IEEE~754. String literals (\texttt{"alice"}) map to \texttt{symbol} through a global symbol table, giving the runtime dense integer identifiers while preserving source-level readability; \texttt{domain} declarations introduce named aliases. Type consistency is enforced by predicate schemas and lowering-time checks: variable occurrences must agree with declared signatures, constants are validated against their destination columns, and arithmetic expressions preserve scalar types.

\lstinputlisting[style=xlogstyle,
                 caption={A well-typed xlog program: typed bridge predicate.},
                 label={lst:types-ok}]{examples/02_types_ok.xlog}

Declaring \texttt{bridge} so that its head argument is constrained at two incompatible types (e.g.\ \texttt{symbol} in the head but \texttt{u32} through \texttt{connected}) rejects the program at compile time.

\subsection{Expressions, Arithmetic, and User-Defined Functions}\label{sec:lang:arith}

Rule bodies may contain arithmetic, comparisons, and calls. Arithmetic bindings use the \texttt{is} operator (\texttt{D is X + Y}); the right-hand side supports \texttt{+ - * / \%}, the built-ins \texttt{abs}, \texttt{min}, \texttt{max}, \texttt{pow}, \texttt{cast}, and user-defined functions. Comparisons compile to \texttt{Filter} nodes and push below joins where profitable. Float comparison uses a total ordering---\texttt{NaN} and signed zeros yield deterministic results rather than contaminating downstream aggregations.

\lstinputlisting[style=xlogstyle,
                 caption={Arithmetic in rule bodies: pairwise proximity via squared Euclidean distance.},
                 label={lst:arith}]{examples/03_arith.xlog}

A user-defined function (UDF) is introduced with \texttt{func}, with optional parameter types and a return type after \texttt{->}. Arithmetic bodies (optionally with \texttt{if/then/else}) inline into the same \texttt{Expr} trees as built-in arithmetic and participate in the same predicate-pushdown and expression-level optimization. UDFs are pure scalar functions and do not recurse through relations, so a UDF call in a recursive rule does not enlarge the derivable set---closing part of the gap to classical Datalog without breaking monotonicity.

\lstinputlisting[style=xlogstyle,
                 caption={User-defined function: \texttt{clamp01} normalizes a raw score into the unit interval.},
                 label={lst:udf}]{examples/04_udf.xlog}

\subsection{Modules and Imports}\label{sec:lang:modules}

Programs decompose into \texttt{.xlog} modules on a search path. The \texttt{use} directive loads a module (\texttt{use graph.}) or selectively imports names (\texttt{use utils/math::\{abs, clamp\}.}); a \texttt{private} modifier hides declarations from importers. Name resolution runs before type inference: the resolver traverses the import graph, detects cycles, merges symbol tables, and surfaces a single logical program, so imported rules participate in exactly the same SCC analysis and GPU semi-naive evaluation as if written inline.

\lstinputlisting[style=xlogstyle,
                 caption={Module \texttt{graph\_lib.xlog}: predicate declarations and recursive reach rules.},
                 label={lst:graph-lib}]{examples/graph_lib.xlog}

\lstinputlisting[style=xlogstyle,
                 caption={Entry \texttt{graph\_main.xlog}: imports \texttt{graph\_lib} and queries \texttt{reach}.},
                 label={lst:graph-main}]{examples/graph_main.xlog}

\subsection{Aggregations and Constraints}\label{sec:lang:agg}

Aggregation operators---\texttt{count}, \texttt{sum}, \texttt{min}, \texttt{max}, \texttt{logsumexp}---appear at rule heads with an explicit aggregation variable and lower to \texttt{GroupBy} nodes executed as radix-sort-and-reduce kernels; the stratifier rejects aggregation inside a recursive SCC. Integrity constraints are headless rules (\texttt{:- body.}) asserting that \texttt{body} is unsatisfiable once evaluation reaches fixpoint; a violation raises a diagnostic. \texttt{logsumexp} is included because it is the workhorse aggregation for probabilistic circuits, where log-space numerical stability is essential (\Cref{sec:prob}).

\lstinputlisting[style=xlogstyle,
                 caption={Stratified aggregation: per-source out-degree with an integrity constraint.},
                 label={lst:agg}]{examples/06_agg.xlog}

\subsection{Completeness over Finite Domains}\label{sec:lang:completeness}

The surface extends toward language completeness while preserving GPU-native execution: every construct normalizes into the existing typed AST, RIR, and runtime paths, and forms that cannot be lowered safely are rejected at compile time rather than executed on a hidden CPU interpreter. The accepted extensions are finite typed lists (\texttt{[]}, \texttt{[H|T]}, the \texttt{list<T>} column type), statically-resolvable meta-predicates (\texttt{ground}, \texttt{functor}, \texttt{=..}, source-fact \texttt{findall}, \texttt{maplist}), explicit stratified negation-as-failure over deterministic relations, magic-set rewriting of bound recursive queries, and finite probabilistic aggregates---each lowering to ordinary relational joins and aggregations. Open-ended generators, dynamic database mutation, unrestricted \texttt{call/N}, and runtime-variable predicate names remain outside the contract.
