\section{GPU-Native Symbolic Substrate}\label{sec:datalog}

The deterministic Datalog engine is the device-resident core the neural bridge and knowledge compiler build on: it is what keeps the symbolic half of a neurosymbolic program on the GPU. The algorithmic approach---semi-naive fixpoint iteration over stratified programs---is well established; the contribution here is engineering. Every relational operator runs as a custom CUDA kernel, delta relations are maintained on-device, and no host--device transfers occur during evaluation.

\subsection{Semi-Naive Evaluation on GPU}

The executor processes an execution plan as strata ordered by the dependency analysis of \Cref{sec:arch}. Non-recursive SCCs evaluate each rule once, dispatching its RIR tree to the appropriate kernel and merging results per head predicate. Recursive SCCs run semi-naive iteration (\Cref{alg:seminaive}): each rule is re-evaluated through delta-rewritten variants---one per recursive scan occurrence, so a self-join such as \texttt{p(X,Y), p(Y,Z)} contributes two variants---and the new tuples are unioned, differenced against the full relation, and deduplicated, all on-device, until no predicate produces new tuples. The profiler records per-operator timing, row counts, and peak memory, feeding the optimizer.

\begin{algorithm}[htbp]
\caption{GPU semi-naive evaluation of a recursive SCC. All buffers are device-resident.}
\label{alg:seminaive}
\begin{algorithmic}[1]
\Require SCC rules $\mathcal{R}$; relation store $S$ (device-resident)
\Ensure least fixpoint of $\mathcal{R}$ merged into $S$
\For{each rule $r \in \mathcal{R}$}
  \State $\Delta_r \gets \textsc{Eval}(r, S) \setminus S$ \Comment{seed delta; set difference on device}
\EndFor
\Repeat
  \For{each rule $r \in \mathcal{R}$ and each recursive scan $i$ in $r$}
    \State $T_{r,i} \gets \textsc{Eval}(r[\Delta \text{ at scan } i], S)$ \Comment{one delta-variant per self-join occurrence}
  \EndFor
  \For{each head predicate $p$}
    \State $\Delta_p \gets \big(\textstyle\bigcup_{r,i \text{ for } p} T_{r,i}\big) \setminus S_p$; \quad dedup $\Delta_p$
    \State $S_p \gets S_p \cup \Delta_p$ \Comment{union + sorted dedup on device}
  \EndFor
\Until{all $\Delta_p = \emptyset$ \textbf{or} iteration cap reached}
\State free all $\Delta$ buffers
\end{algorithmic}
\end{algorithm}

\subsection{Relational Kernels}

xlog implements its relational algebra as direct CUDA kernels rather than wrappers over cuBLAS, cuSPARSE, or Thrust. A library operator imposes its own allocation, synchronization, and host-staging behavior, which would defeat the zero-transfer guarantee of \Cref{sec:arch}; bespoke kernels keep every intermediate buffer on-device. \Cref{tab:kernels} summarizes them.

\begin{table}[htbp]
  \centering
  \caption{Core relational CUDA kernels.}
  \label{tab:kernels}
  \small
  \begin{tabularx}{\linewidth}{@{}l L@{}}
    \toprule
    \textbf{Kernel} & \textbf{Purpose} \\
    \midrule
    \texttt{join}    & Hash join with linked-list collision chains and FNV-1a composite hashing over raw key bytes, so all scalar types share one path; count\,$\to$\,scan\,$\to$\,materialize variants for inner and left-outer joins. \\
    \texttt{sort}    & Stable 4-bit radix sort (8 passes over 32-bit keys): per-pass histogram, prefix sum, scatter. \\
    \texttt{filter}  & Comparison operators for column-vs-constant and column-vs-column, with stream compaction via prefix sum. \\
    \texttt{groupby} & Sorted-input grouped aggregation: boundary detection, then per-group Count/Sum/Min/Max/LogSumExp reduction. \\
    \texttt{dedup}   & Sort-based deduplication and set difference with type-aware equality, including IEEE~754 float handling. \\
    \texttt{set\_ops} & Union (concat\,+\,sort\,+\,dedup) and set difference via sorted-array binary search. \\
    \texttt{scan}    & Blelloch parallel prefix sum---the building block for filter compaction, dedup, and group-by. \\
    \texttt{pack}    & Multi-column key packing into row-major byte arrays with FNV-1a hashing. \\
    \bottomrule
  \end{tabularx}
\end{table}

The filter kernel makes a correctness-critical choice for floating point. Equality uses standard IEEE~754 semantics ($\mathrm{NaN}\neq\mathrm{NaN}$); ordered comparisons use a total-ordering transform that maps an \texttt{f64} bit pattern to an \texttt{i64} whose integer order matches the IEEE~754 \texttt{totalOrder} predicate ($-\mathrm{NaN} < -\mathrm{Inf} < $ negatives $< -0.0 < +0.0 <$ positives $< +\mathrm{Inf} < +\mathrm{NaN}$), so Datalog programs produce deterministic results for all floating-point inputs.

\subsection{Adaptive Join Planning}

The optimizer performs cost-based transformations using runtime statistics. Each plan node is costed along four dimensions---expected rows, coordination cost, peak GPU memory, and host--device transfers---with transfers penalized heavily so the planner prefers device-resident plans even at higher raw compute cost. Predicate pushdown decomposes filters above joins into left-, right-, and cross-predicates; join ordering uses dynamic programming up to ten relations and a greedy heuristic beyond, with selectivities learned from prior executions. The executor can export a statistics snapshot and merge it back before the next compilation, closing the loop between execution and optimization.

\subsection{Worst-Case Optimal Joins}

Binary-join plans can materialize intermediates asymptotically larger than the final result---the classic pathology for cyclic queries such as triangle enumeration. This is not a corner case for xlog's target workloads: program-analysis queries (points-to, call-graph and reachability inference over code-dependency graphs) are dominated by cyclic, skewed joins where the intermediate blowup governs cost. xlog addresses this with a worst-case-optimal join (WCOJ) subsystem~\cite{ngo2018wcoj,veldhuizen2014leapfrog,wang2023freejoin}. A pure-Rust hypergraph planner analyzes rule bodies for multiway-join eligibility, derives a deterministic variable ordering from cardinality and selectivity statistics, and emits a \texttt{MultiWayJoin}/\texttt{ChainJoin} node when promotion provably preserves output semantics; otherwise the binary-join plan is retained as a certified fallback. The GPU kernel executes the join via count\,$\to$\,device prefix scan\,$\to$\,materialize over flat sorted-column storage, with a four-byte metadata D2H total and no count-vector transfer. Coverage spans the triangle, 4-cycles, $K$-clique templates ($K{=}5$--$8$), recursive/SCC integration, and helper-relation splitting that exposes inner skew to root-level histograms. A default-on skew classifier decides per query whether to dispatch WCOJ: high-skew inputs take the WCOJ path, while uniform or empty inputs fall back to the binary-join chain. \Cref{sec:eval} reports the measured speedup.

\subsection{Runtime Optimization}

Interactive and long-running deployments---REPL sessions, iterative solver loops, incremental training---repeatedly evaluate similar queries over relations that change little between calls. A stateless executor rebuilds join indices and recomputes shared subplans each time, paying the dominant cost repeatedly. Three optimizations exploit this between-call stability, each with explicit controls and zero added data-plane transfers: common-subexpression elimination caches safe deterministic subplans within an evaluation; adaptive re-optimization adopts a compiler-supplied candidate plan when mis-plan telemetry crosses a configurable ratio, with output-equivalence checks and rollback; and a persistent hash-index manager---keyed on relation generation, schema, and device---retains built indices across repeated sessions with deterministic LRU eviction (\Cref{sec:eval}).

\subsection{Reversible Symbols}

Datalog programs operate on strings, but GPU kernels operate on fixed-width numeric columns. xlog interns each unique string to a dense, sequential \texttt{u32} identifier and resolves it back in $O(1)$ at output time. Symbol columns then \emph{are} ordinary \texttt{u32} columns---join, sort, filter, and dedup kernels operate on them with no special-casing---and the string table stays host-side, since variable-length data is needed only at ingestion and output. The \texttt{symbol} type maps to Arrow's \texttt{Dictionary(UInt32, Utf8)}, so export to columnar consumers (cuDF, Polars, DuckDB) is zero-copy while preserving the compact internal representation.
