\section{Verified Knowledge Compilation}\label{sec:prob}

Knowledge compilation is the bridge that makes a neural network and a logic program differentiable end-to-end. A network's softmax becomes a distribution over weighted facts; provenance tracking over the program yields a weighted Boolean formula; and that formula compiles into an arithmetic circuit whose forward pass is exact weighted model counting (the query probability) and whose backward pass yields exact gradients with respect to the weights---hence with respect to the network outputs. Probabilistic systems such as ProbLog~\cite{problog2} follow this compile-once/evaluate-many model on the CPU, so a neural pairing like DeepProbLog~\cite{deepproblog} ping-pongs circuit values and gradients across the PCIe bus every iteration. xlog runs the entire pipeline on the GPU---and, crucially, \emph{proves the compiled circuit correct on-device}, so the bridge is sound rather than assumed.

\subsection{The Compilation Pipeline}

The pipeline transforms a probabilistic program into a GPU-resident arithmetic circuit (the XGCF format) through five stages (\Cref{fig:kc}); each is one or more CUDA kernel launches, with all intermediate state device-resident.

\begin{figure*}[tbp]
  \centering
  \begin{tikzpicture}[x=1mm,y=1mm,
      stage/.style={fproc,anchor=north west,text width=24mm,
                    inner xsep=2mm,inner ysep=1.8mm,minimum height=14.5mm}]
    % ---- forward pipeline ----
    \node[stage] (s1) at (0,0)
      {provenance $\varphi$\\[0.4mm]\fsubt{hash-consing}};
    \node[stage] (s2) at (34,0)
      {CNF $\Phi$ (Tseitin)\\[0.4mm]\fsubt{prefix-sum var IDs}};
    \node[stage] (s3) at (68,0)
      {d-DNNF (D4)\\[0.4mm]\fsubt{BFS + block DFS}};
    \node[stage] (s4) at (102,0)
      {XGCF circuit $C$\\[0.4mm]\fsubt{levelized DAG}};
    \node[stage] (s5) at (142,0)
      {evaluate\\[0.4mm]\fsubt{forward $\log Z$}\\[-0.2mm]\fsubt{backward $\nabla$}};
    \node[diamond,aspect=1.2,draw=figaccent,fill=white,line width=0.8pt,
          inner sep=0.5pt] (gate) at (136,-7.25) {\scriptsize\checkmark};
    \draw[farrow] (s1) -- (s2);
    \draw[farrow] (s2) -- (s3);
    \draw[farrow] (s3) -- (s4);
    \draw[farrow] (s4) -- (gate);
    \draw[farrow] (gate) -- (s5);
    % ---- verification lane ----
    \node[fcert,anchor=north,text width=78mm,inner ysep=1.8mm] (cert) at (82,-26)
      {on-GPU CDCL equivalence certificate\\[0.4mm]
       $\varphi\wedge\lnot C$ and $C\wedge\lnot\varphi$ both \textsc{unsat}
       $\Rightarrow C\equiv\varphi$\\[0.4mm]
       \fsubt{resolution proofs checked on-device}\\[-0.2mm]
       \fsubt{complete proof, not sampling}};
    \draw[fver] (s1.south) |- (cert.west);
    \node[flabel,anchor=west] at (15,-20.5) {$\varphi$};
    \draw[fver] (s4.south) -- (s4.south |- cert.north);
    \node[flabel,anchor=west] at (117,-20.5) {$C$};
    \draw[fver] (cert.east) -| (gate.south);
    \node[flabel,anchor=west] at (137.5,-19) {proven $C\equiv\varphi$};
    \node[fnode,anchor=north,draw=figaccent,font=\scriptsize,
          text=figaccent,inner ysep=1.4mm,minimum height=5mm] (trap) at (82,-49)
      {$\times$ device-side trap};
    \draw[fver] (cert.south) -- (trap.north);
    \node[flabel,anchor=west] at (83.5,-45.8) {otherwise};
    % ---- GPU device plane ----
    \begin{scope}[on background layer]
      \node[fplane,fit=(s1)(s5)(cert)(trap),inner sep=3mm] (plane) {};
    \end{scope}
    \node[fplanetitle,anchor=west] at ([xshift=4mm]plane.north west)
      {\planetitle{gpu device-resident --- all stages}};
  \end{tikzpicture}
  \caption{Verified knowledge-compilation pipeline. Every stage runs as CUDA
  kernels on device-resident state. The compiled circuit $C$ reaches
  evaluation only through the equivalence gate (\checkmark): an on-GPU CDCL
  solver proves $\varphi\wedge\lnot C$ and $C\wedge\lnot\varphi$ both
  unsatisfiable, and a failed proof halts in a device-side trap instead of
  reaching evaluation.}
  \label{fig:kc}
\end{figure*}

\textbf{Provenance extraction.} Provenance over deterministic evaluation produces a weighted Boolean formula~\cite{green2007provenance}. An on-device interner hash-conses node batches through an atomic GPU hash table, deduplicating structurally identical subformulas; this is essential because grounding can produce exponentially many redundant subformulas, and interning on the host would force the uninterned graph into host memory---reintroducing the very transfer the pipeline exists to avoid.

\textbf{CNF encoding.} A Tseitin~\cite{tseitin} transformation lowers the formula to CNF on the GPU, using parallel prefix sums to assign compact, gap-free variable IDs and emit clauses into a device-resident compressed-sparse-row structure.

\textbf{D4 compilation.} The CNF compiles into a Decision-DNNF circuit~\cite{darwiche2001dnnf} via a GPU-native variant of the D4 algorithm~\cite{d4compiler}: a BFS frontier exposes parallelism, per-block DFS workers perform component detection, decision-variable selection by VSIDS-like activity~\cite{chaff}, and recursive decomposition. A smoothing pass forces every branch of an OR/decision node to mention the same variable set, and the result is levelized so each topological level evaluates in one kernel launch. Decomposability, determinism, and smoothness---the structural properties that make weighted model counting over a Decision-DNNF correct~\cite{darwicheMarquis2002,chaviraDarwiche2008}---are thus established by construction of this lowering, not by the equivalence check below.

\textbf{Verification and evaluation} are described next.

\subsection{Verification: a Machine-Checked Certificate}

Rather than trust the compiler, xlog proves the compiled circuit $C$ equivalent to its source formula $\varphi$ (\Cref{alg:verifiedkc}). It solves $\varphi \land \lnot C$ and $C \land \lnot \varphi$ on a GPU CDCL solver; both must be unsatisfiable for $C \equiv \varphi$. This is a complete formal proof, not a sampling check: UNSAT results carry a resolution proof checked on-device, and the solver never reads its status back to the host---a wrong answer triggers a device-side trap rather than silently propagating. This certifies that the circuit computes the same Boolean function as $\varphi$ (the count-correctness invariants are guaranteed separately, by construction, as noted above); the compilation back-end thus becomes a machine-checked stage rather than an assumption, closing the failure mode in which a silently miscompiled circuit would corrupt every downstream probability and gradient. The check is on the back-end stage specifically: provenance extraction and CNF encoding are trusted, not certified. CPU-side certified knowledge compilation exists~\cite{capelli2019}; the contribution here is performing the check on-device, sharing the GPU substrate. The same device-resident CDCL substrate is reused for SAT/MaxSAT queries and for epistemic candidate validation (\Cref{sec:epistemic}).

\begin{algorithm}[htbp]
\caption{GPU-native verified knowledge compilation. The circuit is accepted only if proven equivalent to its source formula on-device.}
\label{alg:verifiedkc}
\begin{algorithmic}[1]
\Require weighted Boolean formula $\varphi$ from provenance (device-resident)
\Ensure verified arithmetic circuit $C$ (XGCF), or a device-side trap
\State $\varphi \gets \textsc{InternHashCons}(\varphi)$ \Comment{dedup subformulas via atomic GPU hash table}
\State $\Phi \gets \textsc{TseitinCNF}(\varphi)$ \Comment{prefix-sum variable IDs; CSR clauses}
\State $C \gets \textsc{D4}(\Phi)$ \Comment{BFS frontier + per-block DFS; VSIDS decisions; smooth; levelize}
\State $u_1 \gets \textsc{CDCL}(\varphi \land \lnot C)$;\quad $u_2 \gets \textsc{CDCL}(C \land \lnot \varphi)$
\If{$u_1 \neq \textsc{Unsat}$ \textbf{or} $u_2 \neq \textsc{Unsat}$}
  \State device-side trap \Comment{miscompilation never reaches host or downstream}
\EndIf
\State check resolution proofs of $u_1, u_2$ on-device
\State \Return $C$ \Comment{certificate: $C \equiv \varphi$, machine-checked}
\end{algorithmic}
\end{algorithm}

The verified circuit supports a level-parallel forward pass (log-space weighted model counting, one kernel per level, yielding $\log Z$) and a reverse-order backward pass that accumulates per-variable gradients. Both leave their results in GPU-resident buffers consumable directly by PyTorch via DLPack (\Cref{sec:nesy}).

\subsection{Circuit Caching}

D4 compilation is the most expensive stage---on the order of a minute for a cold start on the MNIST-addition benchmark---and a naive implementation pays it every epoch. The key observation is that the circuit \emph{structure} depends on the propositional structure of the grounded program and the evidence pattern, but \emph{not} on the numerical weights. When neural parameters change between iterations, the weights on probabilistic facts change while the Boolean formula and its compiled circuit stay identical. The compiled circuit is therefore cached, keyed by a structural hash of the provenance graph and evidence configuration; subsequent iterations update only the weight buffers and run the forward/backward kernels. On MNIST addition this amortization yields a $2.74\times$ end-to-end training speedup (\Cref{sec:eval}).

\subsection{Monte Carlo Fallback}

When exact compilation is infeasible---the ground formula is too large for D4 within the GPU memory budget, or the user accepts approximate results---xlog provides a GPU-parallel Monte Carlo engine. It draws values for all probabilistic facts on the GPU and evaluates the deterministic program in each sampled world, using rejection sampling or, when every evidence atom is a Bernoulli fact, evidence clamping that forces observed values and eliminates rejection waste. Both methods report confidence intervals. The sampler is a GPU-resident megakernel that evaluates each world in a single device launch with a bounded device-side fixpoint; unsupported fragments are rejected fail-closed with a typed diagnostic rather than silently falling back to the host.
