\section{Neurosymbolic Integration}\label{sec:nesy}

This section is where the three pillars meet. The device-resident substrate (\Cref{sec:datalog}) keeps the symbolic state on the GPU; verified knowledge compilation (\Cref{sec:prob}) turns a network's outputs into a sound differentiable circuit; and zero-copy coupling carries gradients back into the neural optimizer without leaving the device. Together they let neural perception and logical reasoning train jointly, end to end, with no host--device transfers in the loop.

\subsection{Neural Predicates}\label{subsec:neural-pred}

In xlog a neural network \emph{is} a predicate. The \texttt{nn/4} declaration embeds a network directly into the program:

\begin{lstlisting}[style=xlogstyle,
                   caption={Neural predicate declaration (\texttt{nn/4}).},
                   label={lst:nn-decl}]
nn(network_name, [input_vars], output_var,
   [output_labels]) :: predicate(args).
\end{lstlisting}

This is not a foreign-function escape hatch: it states that evaluating \texttt{predicate(args)} requires a forward pass through the named network, whose softmax over the label set becomes a distribution over probabilistic facts feeding the knowledge-compilation pipeline of \Cref{sec:prob}. On the Python side a PyTorch module is registered against the predicate:

\begin{lstlisting}[style=pythonstyle,
                   caption={Registering a PyTorch module against a neural predicate.},
                   label={lst:nn-register}]
program.register_network("mnist_net", net, optimizer)
\end{lstlisting}

The dataflow runs in two directions. \textbf{Forward:} when evaluation reaches an \texttt{nn/4}-backed predicate, the registered module runs, and its softmax vector is decomposed into weighted probabilistic facts---one per label---fed into the PIR/CNF/D4/XGCF pipeline. Because circuit structure depends on the program, not the weights, the compiled circuit is cached and only the leaf weights are updated on later iterations. \textbf{Backward:} after the circuit computes the forward log-probability and backward gradients via level-parallel kernels, the per-leaf gradient buffers are exported as DLPack tensors and injected into PyTorch's autograd graph, and the optimizer updates the network as usual. The GIL is released during GPU work, so CUDA execution does not block the interpreter. A neural predicate is not confined to a query head, moreover: it may also appear inside rule bodies, so that whole rules---not just a single perceptual predicate---become the trainable object (\Cref{subsec:trainable-bodies}).

\subsection{End-to-End Training}

The canonical example is MNIST addition (the DeepProbLog benchmark): given two digit images, predict their sum, with supervision only on sums---the network never sees individual digit labels. The entire reasoning structure is two rules:

\begin{lstlisting}[style=xlogstyle,
                   caption={MNIST addition: a CNN-backed digit predicate and an addition rule.},
                   label={lst:mnist-prog}]
nn(mnist_net, [X], Y, [0,1,2,3,4,5,6,7,8,9])
    :: digit(X, Y).
addition(X, Y, Z) :-
    digit(X, D1), digit(Y, D2), Z is D1 + D2.
\end{lstlisting}

The \texttt{nn/4} declaration connects the CNN to \texttt{digit/2}; the addition rule computes every consistent decomposition of a sum through probabilistic marginalization. Each query \texttt{addition(i, j, s)} trains by minimizing $-\log P(\texttt{addition}(i, j, s))$ under the compiled circuit. The driver compiles the program, registers the network, and runs the loop, which batches queries sharing a circuit template and runs forward/backward over the cached circuit:

\begin{lstlisting}[style=pythonstyle,
                   caption={Python driver: compile, register, train.},
                   label={lst:mnist-train}]
program = pyxlog.Program.compile(SRC)
program.register_network("mnist_net", net, optimizer)
program.add_tensor_source("train", train_images)
history = pyxlog.train_model_tensor(
    program, queries,
    epochs=epochs, batch_size=batch_size)
\end{lstlisting}

\begin{figure*}[tbp]
  \centering
  \begin{tikzpicture}[x=1mm,y=1mm,
      ring/.style={anchor=north west,text width=26mm,
                   inner xsep=2mm,inner ysep=1.8mm,minimum height=14.5mm}]
    % ---- forward row ----
    \node[fneural,ring] (net) at (0,0)
      {neural network\\[0.4mm]\fsubt{PyTorch forward}};
    \node[fproc,ring] (ground) at (54,0)
      {symbolic grounding\\[0.4mm]\fsubt{rules $\cdot$ provenance}};
    \node[fproc,ring] (circ) at (96,0)
      {verified circuit\\[0.4mm]\fsubt{proven once $\cdot$ cached}};
    \node[fneural,ring] (loss) at (138,0)
      {loss $-\log P$\\[0.4mm]\fsubt{per-query NLL}};
    \node[circle,draw=figaccent,fill=white,line width=0.7pt,inner sep=0.4pt,
          font=\tiny] at (circ.north west) {\checkmark};
    % ---- backward row ----
    \node[fproc,ring] (back) at (138,-27)
      {circuit backward\\[0.4mm]\fsubt{level-parallel adjoint}};
    \node[fneural,ring] (opt) at (0,-27)
      {optimizer step\\[0.4mm]\fsubt{PyTorch autograd}};
    % ---- ring edges ----
    \draw[farrow] (net) -- node[flabel,above=0.4mm,text width=21mm]
      {softmax $\to$ weighted facts} (ground);
    \draw[farrow] (ground) -- (circ);
    \draw[farrow] (circ) -- node[flabel,above=0.4mm,text width=11mm]
      {forward $\log P$} (loss);
    \draw[fback] (loss) -- node[flabel,anchor=east,xshift=-1mm] {backward} (back);
    \draw[fback] (back) -- node[flabel,above=0.4mm]
      {per-leaf gradients exported via DLPack (zero-copy)} (opt);
    \draw[fback] (opt) -- node[flabel,anchor=west,xshift=1mm]
      {update $\theta$} (net);
    \node[flabel] at (84,-20)
      {one training iteration --- circuit \& proof reused across iterations};
    % ---- GPU device plane around the whole loop ----
    \node[flabel] (foot) at (84,-45.5)
      {zero host--device transfers in the loop (byte-accounted)};
    \begin{scope}[on background layer]
      \node[fplane,fit=(net)(loss)(back)(opt)(foot),inner sep=3mm] (plane) {};
    \end{scope}
    \node[fplanetitle,anchor=west] at ([xshift=4mm]plane.north west)
      {\planetitle{gpu device-resident}};
    % ---- legend ----
    \draw[figgreen,line width=0.9pt,fill=white,rounded corners=1pt]
      (38,-56.5) rectangle +(3,3);
    \node[fsub,anchor=west] at (42.5,-55) {PyTorch};
    \draw[figblue,line width=0.9pt,fill=white,rounded corners=1pt]
      (62,-56.5) rectangle +(3,3);
    \node[fsub,anchor=west] at (66.5,-55) {xlog substrate};
    \draw[figaccent,line width=0.9pt,fill=white,rounded corners=1pt]
      (96,-56.5) rectangle +(3,3);
    \node[fsub,anchor=west] at (100.5,-55) {verified (\Cref{fig:kc})};
  \end{tikzpicture}
  \caption{One end-to-end neurosymbolic training iteration. The entire
  loop---neural forward pass, symbolic grounding, circuit evaluation, loss,
  and gradient return---executes on one CUDA device (shaded plane) with zero
  host--device data-plane transfers. The circuit is compiled and proven
  correct once (\checkmark, \Cref{fig:kc}), then reused across iterations,
  and gradients re-enter PyTorch's autograd graph zero-copy through DLPack.}
  \label{fig:training}
\end{figure*}

The decisive performance property (\Cref{fig:training}) is that compilation and verification happen once, on the first forward pass; every later iteration reuses the cached circuit, executing only the weight update and the level-parallel forward/backward kernels. This is the mechanism behind the $2.74\times$ training speedup; the loss reduction across seeds confirms that gradients flow from circuit evaluation through the logic program back to the CNN, and the full timing breakdown is reported in \Cref{sec:eval}.

\subsection{Zero-Copy Interoperability}\label{subsec:interop}

A GPU-native engine is only useful if results reach the frameworks researchers already use---without copies that would reintroduce the transfer wall. xlog exposes its device-resident state through four standard interfaces.

\textbf{DLPack.} Each relation column or gradient buffer becomes a managed DLPack~\cite{dlpack} tensor whose GPU memory is reference-counted and freed only when every consumer releases it; on import, xlog validates device, dtype, contiguity, and alignment, optionally against an expected schema. In Python the protocol surfaces as a capsule, so \texttt{torch.from\_dlpack} yields a live CUDA tensor backed by the very memory xlog wrote.

\textbf{Arrow.} For columnar tools (Polars, DuckDB, cuDF), relations export through the Arrow~\cite{arrow} C Device Data Interface; symbol columns export as Arrow dictionary arrays, so a consumer reads symbolic columns as native string dictionaries while xlog keeps the compact integer representation.

\textbf{Python bindings.} The \texttt{pyxlog} extension (PyO3~\cite{pyo3}) exposes compilation, evaluation, training, and result extraction; tensor-valued results are returned as DLPack capsules, and long-running GPU operations release the GIL so data-loader threads proceed concurrently.

\textbf{PyTorch autograd.} The circuit behaves as a differentiable function inside PyTorch: the network's grad-enabled forward feeds the circuit, the scalar loss is exported via DLPack and re-imported, and a batched path stacks inputs into one forward pass and runs a single backward---so gradients flow from the logic layer back to \texttt{nn.Module} parameters with no custom autograd \texttt{Function}.

\subsection{Trainable Rule Bodies}\label{subsec:trainable-bodies}

Placing a neural predicate in a rule body lifts xlog from learning a single perceptual predicate to learning whole rules, while the pipeline stays device-resident with no host round-trips.

\textbf{Mixed bodies.} A trainable body may interleave neural-evaluated atoms with ordinary symbolic literals. The symbolic literals act as hard join conditions: they gate which groundings can fire but carry no probability mass or gradient, so the head probability factors as the (satisfiable?) symbolic condition times the neural probability scaled by a learned per-rule weight~$\sigma(w)$, and gradients flow only through the network outputs and the rule weight, never through the ground facts.

\textbf{Existential joins.} When an existential (non-head) body variable is a neural predicate's input, xlog grounds that predicate over the \emph{real} join domain \emph{inside} the circuit rather than stripping the join as a pre-filter: the neural occurrence expands into one circuit leaf per domain event, and provenance OR-aggregates $\exists\,e.\ \mathrm{neural}(e)\wedge\mathrm{join}(e,\mathit{head})$ at each head binding. This makes ``does supporting evidence exist?'' rules learnable---reasoning over a whole domain of events---rather than only per-instance classification.

\textbf{Joint multi-rule mixtures.} Several same-head trainable clauses combine as a noisy-OR mixture, $P(\mathit{head}_i)=1-\prod_k\bigl(1-r_k[i]\,m_k[i]\,\sigma(w_k)\bigr)$, where $r_k[i]$ is a candidate's relational eligibility, $\sigma(w_k)$ its learned rule weight, and $m_k[i]$ an optional graded confidence in $[0,1]$ carried per binding (for example, a fact's confidence trajectory over time). Rule learning thus moves from binary rule eligibility to a continuous, differentiable weighting of competing and cooperating rules.

These trainable-rule paths share the verified circuit and the zero-copy substrate of the simpler predicate case, and extend to recursive programs now that provenance converges on two-sided recursive strongly-connected components. Like the differentiable-ILP path below, the surface is validated by demonstrations rather than head-to-head benchmarks and is a beta capability (\Cref{sec:limits}).

\subsection{Term Embeddings and Differentiable ILP}

Two further paths ride the same device-resident bridge. \textbf{Term embeddings} attach dense, optionally trainable vectors to logical symbols: where \texttt{nn/4} maps perception to discrete facts, embeddings give logical entities a continuous representation that participates in neural computation, with gradients flowing through the embedding lookup---enabling, for example, knowledge-graph entity representations trained jointly with neural perception. \textbf{Differentiable ILP} learns first-order rules rather than network weights: candidate rules are represented as sparse bitmasks over the ground-atom space, credit assignment runs entirely on-device via scatter--gather, and a six-gate promotion pipeline (convergence, novel-rate, regression, held-out F1, ambiguity, typed-schema checks) controls which rules are committed. Where the trainable rule bodies of \Cref{subsec:trainable-bodies} learn rule \emph{weights} and per-binding confidences, differentiable ILP learns rule \emph{structure}; the two are complementary. This sparse, GPU-resident formulation avoids the cubic blowup of dense rule materialization; it is a beta subsystem (\Cref{sec:limits}).
