\section{Introduction}\label{sec:intro}

Symbolic AI and neural AI have followed divergent engineering paths. Datalog engines, probabilistic logic systems such as ProbLog, and inductive logic programming (ILP) frameworks are implemented as CPU-bound interpreters or compilers, processing relations and proofs in main memory. Deep learning frameworks---PyTorch, JAX---execute dense tensor computations on GPUs via highly optimized CUDA kernels. When researchers combine the two paradigms, as in DeepProbLog~\cite{deepproblog} or NeurASP~\cite{neurasp}, the symbolic component remains on the CPU while the neural component runs on the GPU. Every training iteration transfers data across the PCIe bus: the CPU-side logic engine materializes query results, ships them to the GPU for gradient computation, then pulls gradients back to update symbolic parameters. At scale---millions of ground atoms, thousands of training steps---these host--device transfers dominate wall-clock time and memory bandwidth, becoming the primary bottleneck rather than the inference or learning computation itself.

The gap is architectural. Existing systems address individual reasoning tasks on the GPU in isolation: GPU-accelerated Datalog evaluation~\cite{gpulog,vflog}, GPU SAT solvers~\cite{parafrost}, or differentiable logic on CPU with GPU-side neural networks~\cite{dilp}. No single system performs deterministic Datalog evaluation, probabilistic inference via knowledge compilation, SAT/MaxSAT verification, and differentiable neural-symbolic training entirely on the GPU with zero host--device data transfers in production paths. The absence of such a platform forces practitioners into multi-system pipelines---a Datalog engine for rule evaluation, a separate probabilistic reasoner, a Python training loop bridging CPU logic to GPU tensors---each with its own data format, memory model, and failure modes.

xlog addresses this gap as a GPU-native logic programming language whose compiler and runtime span five reasoning paradigms: deterministic logic (semi-naive evaluation with stratified negation), probabilistic inference (exact knowledge compilation and Monte Carlo sampling), epistemic reasoning (world-view semantics over multiple stable models), SAT/MaxSAT solving (GPU CDCL with proof certificates), and neural-symbolic learning (differentiable training with PyTorch interoperability). The system is implemented in Rust with 23 custom CUDA kernel files (17.7K lines of device code) organized into a layered crate architecture. The compilation pipeline transforms Datalog source into a relational intermediate representation (RIR), lowers probabilistic programs through a propositional intermediate representation (PIR) into CNF, compiles decision-DNNF circuits via D4~\cite{d4compiler}, and encodes the result in a GPU-resident circuit format (XGCF) for forward and backward evaluation. All semantic data structures---fact stores, circuit nodes, solver state, gradient buffers---remain GPU-resident during execution. Host involvement is limited to orchestration, I/O, and compilation.

The principal contributions of this paper are:

\begin{itemize}
  \item \textbf{A typed logic programming language for GPU execution.} The \texttt{xlog-logic} crate implements a statically-typed logic language with user-defined functions, modules and imports, stratified aggregation, integrity constraints, probabilistic facts, and neural predicate declarations---all compiling through a single PEG parser, type inferencer, and SCC-based stratifier into the relational IR that feeds every downstream backend (\Cref{sec:lang}). The surface is completeness-oriented over finite domains: finite lists, safe meta-predicates, explicit negation-as-failure, magic-set query planning, and probabilistic aggregates all normalize into the same typed relational paths rather than a CPU term interpreter.
  \item \textbf{GPU-resident semi-naive Datalog evaluation.} The \texttt{xlog-runtime} and \texttt{xlog-cuda} crates implement relational algebra operators (hash join, radix sort, filter, deduplication, set difference, grouped aggregation) as CUDA kernels, executing fixed-point iteration entirely on the GPU with columnar storage and HISA indexing~\cite{gpulog}.
  \item \textbf{Worst-case-optimal multiway joins with cost-aware planning.} A hypergraph planner promotes eligible multiway rules to a GPU worst-case-optimal join (WCOJ) subsystem with cardinality/selectivity-aware variable ordering, recursive-SCC integration, and adaptive skew-triggered dispatch, with deterministic binary-join fallback preserved. On skewed paper-scale program-analysis workloads this yields a measured $27.96\times$ geometric-mean speedup over the binary-join baseline (\Cref{sec:eval}). A runtime optimization layer adds persistent hash indexing, common-subexpression elimination, and adaptive re-optimization for repeated sessions.
  \item \textbf{GPU-native knowledge compilation with verified circuits.} The \texttt{xlog-prob} crate compiles probabilistic Datalog programs through a PIR-to-CNF-to-D4-to-XGCF pipeline, producing GPU-resident arithmetic circuits with compile-once/evaluate-many semantics; forward and backward passes run as level-parallel CUDA kernels for exact weighted model counting and gradient computation without host transfers. Crucially, an on-GPU CDCL solver proves each compiled circuit logically equivalent to its source formula, so circuit correctness is a machine-checked certificate rather than an assumption. To our knowledge this is the first end-to-end GPU-resident knowledge-compilation pipeline with on-device equivalence verification, and it is the paper's principal contribution.
  \item \textbf{GPU-native epistemic reasoning (release candidate).} The v0.9.0 release candidate adds epistemic logic programming over world views: modal \texttt{know}/\texttt{possible} literals are preserved in an Epistemic IR and executed by a Generate--Propagate--Test plan whose phases run as CUDA kernels, with no CPU fallback on the accepted path. It reuses the same on-GPU CDCL solver and worst-case-optimal join machinery as the deterministic and probabilistic backends (\Cref{sec:epistemic}).
  \item \textbf{End-to-end differentiable neural-symbolic training.} The \texttt{xlog-neural} crate and \texttt{pyxlog} Python package connect compiled circuits to PyTorch's autograd graph. Circuit structure depends on the logic program, not on neural network weights, so compiled XGCF templates are cached across training iterations. This circuit caching yields a measured $2.74\times$ end-to-end training speedup (95\% CI: $[2.29, 3.18]$) by eliminating redundant D4 recompilation.
  \item \textbf{Zero-copy interoperability with ML frameworks.} The \texttt{xlog-cuda} crate exposes GPU-resident query results and gradient tensors via DLPack~\cite{dlpack} capsules and Arrow~\cite{arrow} IPC/C Data interfaces, enabling direct consumption by PyTorch, cuDF, and other frameworks without data copies or device synchronization.
  \item \textbf{Differentiable ILP with GPU-resident credit assignment.} The \texttt{pyxlog} dILP trainer implements sparse GPU mask computation, a fully device-resident credit/loss path with zero host transfers, and a six-gate promotion pipeline (convergence, novel-rate audit, regression check, holdout F1 threshold, ambiguity scan, typed schema validation) for transactional rule induction. This subsystem is currently in beta.
\end{itemize}

The remainder of this paper is organized as follows. \Cref{sec:arch} presents the system architecture and crate decomposition. \Cref{sec:lang} describes the xlog language surface. \Cref{sec:datalog} details GPU-native deterministic Datalog evaluation. \Cref{sec:prob} covers the probabilistic inference pipeline. \Cref{sec:nesy} details the neural-symbolic bridge and differentiable training. \Cref{sec:epistemic} introduces epistemic reasoning over world views. \Cref{sec:interop} discusses interoperability with external frameworks. \Cref{sec:eval} presents evaluation results. \Cref{sec:related} surveys related work. \Cref{sec:limits} discusses limitations and future directions.
