\section{The xlog Language}\label{sec:lang}

This section describes the xlog language surface: types, expressions, user-defined functions, modules, and stratified aggregations with constraints. Probabilistic facts (\texttt{p::f.}) and neural predicate declarations (\texttt{nn/k}) are language-level constructs described in \Cref{sec:prob} and \Cref{sec:nesy} respectively.

\subsection{Design Philosophy}\label{sec:lang:philosophy}

xlog is a typed logic programming language designed for GPU execution. Three principles shape its surface:

\textbf{Typed predicates.} Every predicate has a declared signature over a closed set of scalar types (\texttt{u32}, \texttt{u64}, \texttt{i32}, \texttt{i64}, \texttt{f32}, \texttt{f64}, \texttt{bool}, \texttt{symbol}). On current \texttt{main}, predicate schemas and lowering-time validation catch argument mismatches before any GPU kernel is generated; richer standalone type inference exists in reserved frontend APIs but is not yet the universal production entry point.

\textbf{Stratified semantics as contract.} Negation, aggregation, and recursion interact through a strongly-connected-component analysis on the predicate dependency graph. Programs whose SCCs cross negation or aggregation edges are rejected at compile time. Every accepted program has a well-defined fixpoint and a deterministic stratum order, which the runtime relies on for GPU scheduling.

\textbf{One language for four paradigms.} Probabilistic facts (\texttt{p::f.}), annotated disjunctions, neural predicate declarations (\texttt{nn/k}), and SAT constraints share the same syntactic core with deterministic Datalog. Parsing, AST construction, dependency analysis, and most lowering logic are shared across the workspace, while backend-specific expansion and compilation passes refine the resulting plans for deterministic, probabilistic, SAT, and neural-symbolic execution. The distinction between reasoning paradigms lives primarily in the backend, not in the surface syntax.

A minimal xlog program exhibits the core surface:

\lstinputlisting[style=xlogstyle,
                 caption={Minimal xlog program: typed predicates, facts, a recursive rule, a query.},
                 label={lst:hello}]{examples/01_hello.xlog}

Typed predicate declarations, ground facts, recursive rules, and a query appear in eight lines. The remainder of this section expands each component and adds the extensions that move xlog beyond classical Datalog.

\subsection{Type System}\label{sec:lang:types}

Every predicate is declared with a typed signature via \texttt{pred}. Eight scalar types are supported: \texttt{u32}, \texttt{u64}, \texttt{i32}, \texttt{i64} for fixed-width integers; \texttt{f32} and \texttt{f64} for floats with IEEE 754 semantics; \texttt{bool}; and \texttt{symbol} for dictionary-encoded atoms. String literals in source (\texttt{"alice"}) are terms that map to \texttt{symbol} through the \texttt{SymbolTable} in \texttt{xlog-core}, giving the runtime dense integer identifiers while preserving source-level readability. Users can also introduce named type aliases through \texttt{domain} declarations.

Current \texttt{main} enforces type consistency through explicit predicate schemas plus frontend/lowering checks. Variable occurrences across body atoms must agree with declared predicate signatures; literal constants are validated against their destination column types; arithmetic expressions preserve scalar types in the emitted \texttt{Expr} tree. The dedicated pass in \texttt{crates/xlog-logic/src/typeinfer.rs} covers richer inference cases, especially around user-defined functions, but remains a reserved API rather than the sole wired frontend path.

A well-typed program:

\lstinputlisting[style=xlogstyle,
                 caption={A well-typed xlog program: typed bridge predicate.},
                 label={lst:types-ok}]{examples/02_types_ok.xlog}

Changing \texttt{bridge}'s declaration to \texttt{pred bridge(symbol, u32).} would cause inference to report that the head argument \texttt{A} is constrained at \texttt{symbol} (head) and \texttt{u32} (through \texttt{connected}), rejecting the program at compile time.

\subsection{Expressions and Arithmetic}\label{sec:lang:arith}

Rule bodies may contain arithmetic expressions, built-in function calls, and comparison filters. Arithmetic bindings use the \texttt{is} operator following Prolog convention: \texttt{D is X + Y} computes \texttt{X + Y} and binds the result to \texttt{D}. The right-hand side supports the standard arithmetic operators (\texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, \texttt{\%}), the built-ins \texttt{abs}, \texttt{min}, \texttt{max}, \texttt{pow}, and \texttt{cast}, and calls to user-defined functions (\Cref{sec:lang:udf}). Comparison literals (\texttt{<}, \texttt{<=}, \texttt{=}, \texttt{==}, \texttt{!=}, \texttt{>=}, \texttt{>}) compile to \texttt{Filter} RIR nodes and push down below joins where profitable.

Float arithmetic follows IEEE 754 with explicit NaN and $\pm\infty$ handling: \texttt{NaN != NaN}, and ordering predicates place \texttt{NaN} outside any other value rather than raising errors. This matters for aggregations, where a single unguarded NaN would contaminate downstream GPU computations.

A pairwise-proximity query illustrates expression use:

\lstinputlisting[style=xlogstyle,
                 caption={Arithmetic in rule bodies: pairwise proximity via squared Euclidean distance.},
                 label={lst:arith}]{examples/03_arith.xlog}

The rule compiles to a self-join of \texttt{point} with an intermediate \texttt{Project} that materializes \texttt{D}, followed by a \texttt{Filter} on \texttt{D < 1.0}. The optimizer relocates the cheap comparison \texttt{A < B} ahead of the join when profitable, reducing intermediate cardinality.

\subsection{User-Defined Functions}\label{sec:lang:udf}

xlog includes a user-defined function (UDF) surface for pure scalar computations inside rule bodies and expressions. A UDF is introduced with the \texttt{func} keyword; parameters carry optional type annotations, and the return type is declared after a \texttt{->}. Arithmetic-expression bodies (optionally with conditional \texttt{if/then/else}) are the path integrated into the current arithmetic frontend. The syntax also admits a more ambitious predicate-based form \texttt{V :- body.}, but that form remains partially wired and should be viewed as reserved rather than uniformly available across all compiler entry points.

UDFs are intended to be pure: they take scalar arguments and return scalar values, and do not recurse through relations. The function registry in \texttt{crates/xlog-logic/src/function.rs} performs structural validation---duplicate definitions, unresolved calls, name conflicts, and recursive definitions without a guarded base case are rejected. Calls to arithmetic UDFs appear wherever arithmetic expressions do, including as the right-hand side of an \texttt{is}-expression. Expansion is explicit rather than universal on current \texttt{main}: selected execution paths invoke \texttt{crates/xlog-logic/src/expand.rs} to inline arithmetic UDFs into composed \texttt{Expr} trees (\texttt{Add}, \texttt{Mul}, \texttt{Conditional}, and so on), while predicate-bodied expansion remains reserved. Where this expansion path is used, UDFs participate in the same predicate-pushdown and expression-level optimization passes as built-in arithmetic.

Most classical Datalog dialects have no UDF analogue: functors are either absent or restricted to uninterpreted term construction. xlog's arithmetic UDF path closes part of this gap without breaking monotonicity. Because the integrated bodies are pure scalar expressions, a UDF call in a recursive rule does not enlarge the set of derivable facts beyond what the surrounding Datalog rule already derives.

A UDF for score normalization:

\lstinputlisting[style=xlogstyle,
                 caption={User-defined function: \texttt{clamp01} normalizes a raw score into the unit interval.},
                 label={lst:udf}]{examples/04_udf.xlog}

The \texttt{clamp01} UDF lowers to a conditional expression embedded in the \texttt{Project} computing \texttt{norm\_score}'s second column; no host dispatch occurs during evaluation.

\subsection{Modules and Imports}\label{sec:lang:modules}

xlog programs decompose into modules. A module is a \texttt{.xlog} file located on the module search path, which the \texttt{xlog} CLI controls via the \texttt{--module-path} flag. The \texttt{use} directive loads a named module (e.g., \texttt{use graph.}) or selectively imports names from it (\texttt{use utils/math::\{abs, clamp\}.}). Module paths use \texttt{/} as a separator and mirror the on-disk directory structure. Declarations and functions carry an optional \texttt{private} modifier to hide them from importers.

Name resolution runs before type inference. The resolver in \texttt{crates/xlog-logic/src/resolver.rs} (469 lines) traverses the import graph, detects cycles, merges symbol tables, and surfaces a single logical program whose remaining compilation stages proceed as if the user had written all rules inline. Imports are transitive; attempts to reach a \texttt{private} name yield a structured \texttt{ModuleError::PrivatePredicate} diagnostic rather than a silent shadowing.

A two-module program where a top-level entry imports a graph-reachability library:

\lstinputlisting[style=xlogstyle,
                 caption={Module \texttt{graph\_lib.xlog}: predicate declarations and recursive reach rules.},
                 label={lst:graph-lib}]{examples/graph_lib.xlog}

\lstinputlisting[style=xlogstyle,
                 caption={Entry \texttt{graph\_main.xlog}: imports \texttt{graph\_lib} and queries \texttt{reach}.},
                 label={lst:graph-main}]{examples/graph_main.xlog}

The resolver merges the two files before stratification, so the library's recursive \texttt{reach} rules participate in exactly the same SCC analysis and GPU semi-naive evaluation as if they had been written inline.

\subsection{Aggregations and Constraints}\label{sec:lang:agg}

xlog supports stratified aggregation and integrity constraints. Aggregation operators---\texttt{count}, \texttt{sum}, \texttt{min}, \texttt{max}, and \texttt{logsumexp}---appear at head positions of rules, with an explicit aggregation variable. The stratifier rejects any program in which an aggregation participates in a recursive SCC, guaranteeing a well-defined fixpoint order. Aggregations lower to \texttt{GroupBy} RIR nodes, executed on the GPU as radix-sort-and-reduce kernels in \texttt{xlog-cuda}.

Integrity constraints are headless rules: \texttt{:- body.} denotes a program-level assertion that \texttt{body} must be unsatisfiable after evaluation reaches fixpoint. A satisfied constraint is silent; a violated constraint raises a diagnostic. Constraints compose cleanly with stratification---they evaluate after the strata their body predicates inhabit.

A graph-degree computation with an integrity check that every edge source has a degree record:

\lstinputlisting[style=xlogstyle,
                 caption={Stratified aggregation: per-source out-degree with an integrity constraint.},
                 label={lst:agg}]{examples/06_agg.xlog}

The \texttt{degree} rule lowers to a \texttt{GroupBy} over \texttt{edge} with \texttt{count} as the aggregator. The integrity constraint is desugared into an auxiliary rule whose body mirrors the constraint body (so the \texttt{not degree(X, \_)} literal here contributes an \texttt{Anti}-join); at evaluation completion the runtime asserts the auxiliary relation is empty and raises a diagnostic otherwise. \texttt{logsumexp} is included explicitly because it is the workhorse aggregation for probabilistic circuits (\Cref{sec:prob}), where log-space numerical stability is essential.
