\section{Epistemic Reasoning}\label{sec:epistemic}

The reasoning modes described so far ask what is \emph{true}: which atoms a program derives, with what probability, or under what neural evidence. Epistemic logic programming instead asks what is \emph{known} or \emph{possible} across the multiple stable models a program admits---the natural setting for introspection and reasoning under incomplete knowledge, where a query ranges over a program's whole space of world views rather than a single model. xlog adds this mode in its v0.9.x line, executing it on the same GPU-resident substrate as the other backends.

\subsection{Modal Surface and the Epistemic IR}

A program may use four modal literals in rule bodies---\texttt{know}, \texttt{possible}, \texttt{not know}, and \texttt{not possible}---over a predicate. Rather than rewrite these away at parse time, the compiler preserves them in an \emph{Epistemic Intermediate Representation} (EIR) that sits between the AST and GPU lowering, so the world-view semantics are resolved by a dedicated execution plan rather than smuggled into ordinary relational algebra. The accepted surface spans finite nested modal-operator chains (normalized and executed), epistemic integrity constraints (which prune candidate world views on the GPU), rules that jointly couple several epistemic predicates across components, same-name multi-arity modal predicates, and recursive epistemic execution (\Cref{subsec:epistemic-gpu}). The boundaries that remain fail-closed are genuine rather than incidental: raw RIR lowering of modal literals, genuinely cyclic modal coupling with no founded or well-founded order (e.g.\ \texttt{a :- know b.\ b :- know a.}), and unbounded tuple-key forms (cons \texttt{[H|T]}, predicate references, aggregates), each rejected at compile time with a typed diagnostic rather than silently approximated.

\subsection{World-View Semantics}

A world view is a non-empty set of stable models. \texttt{know p} holds when \texttt{p} appears in every model of the world view; \texttt{possible p} holds when it appears in at least one; \texttt{not know} and \texttt{not possible} are their negations. Two semantics are selectable through \texttt{\#pragma epistemic\_mode}. The default, FAEEL (Founded Autoepistemic Equilibrium Logic), rejects circular epistemic support unless it is independently founded, so a self-supporting rule such as \texttt{p :- possible p.} fails closed. A Gelfond-1991-style compatibility mode, \texttt{g91}, instead admits such self-support for legacy epistemic specifications. Non-epistemic programs evaluate identically under both modes.

\subsection{GPU-Native Execution}\label{subsec:epistemic-gpu}

Accepted epistemic programs lower from EIR to a GPU-executable plan rather than to ordinary RIR. Execution follows a Generate--Propagate--Test schedule whose three phases run as CUDA kernels: \emph{generate} enumerates bounded candidate assumptions as device bitsets; \emph{propagate} prunes candidates that immediately contradict known or rejected facts; and \emph{test} validates the survivors against the program's world views, checking stable-model tuple membership on the GPU for predicates of each arity. The reduced ordinary part of each rule reuses the existing optimizer, WCOJ planner, and helper-splitting machinery (\Cref{sec:datalog}), and \emph{epistemic splitting} decomposes a program into independent components that are solved separately and recombined. The accepted hot path performs no CPU fallback---candidate enumeration, membership tests, and world-view validation all execute on device---and a transfer budget tracks that no data-plane host transfers occur within it.

Recursive epistemic programs execute through the same substrate. A modal atom that ranges over an invariant relation reduces to its gated relation and runs through the ordinary GPU recursive fixpoint; a modal over an epistemically \emph{determined} head is resolved by stratified execution (the theorem \(\texttt{know}\,R \equiv R\) applied at the relation-store boundary); positive recursion computes the FAEEL founded least fixpoint, or the G91 self-support fixpoint under \texttt{g91}; and a negated-modal cycle with no stratified order is resolved by a GPU-backed well-founded-semantics (WFS) alternating fixpoint rather than the host probabilistic WFS solver. That WFS path is GPU-backed but still host-orchestrated---it is not yet a device-resident, no-host-interaction WFS, which remains future work (\Cref{sec:limits}).

\subsection{Solver and Probabilistic Integration}

Epistemic candidate validation reuses the on-GPU CDCL solver already built for knowledge-compilation verification (\Cref{sec:prob}): the same device-resident CNF/CDCL substrate handles satisfiability checks, bounded MaxSAT, and assumption-based queries, avoiding a separate epistemic SAT engine. Accepted world-view evidence likewise feeds the existing GPU-native exact/provenance path, so a query can combine epistemic and probabilistic structure without leaving the device. In the v0.9.x line the GPU-native epistemic execution path is the mature surface; the broader solver portfolio and the end-to-end probabilistic--epistemic evidence path are still being completed (\Cref{sec:limits}).
