\section{Evaluation}\label{sec:eval}

This section reports a mix of artifact-backed measurements and benchmark acceptance targets for xlog's deterministic, probabilistic, and neural-symbolic subsystems. Every quantitative claim is tied either to a checked-in artifact or to the benchmark contract in \texttt{docs/BENCHMARKS.md}; the two are distinguished explicitly below. These are single-system measurements on development hardware, not controlled head-to-head comparisons against other engines.

\subsection{Methodology}

\textbf{Hardware.} All measurements were collected on an NVIDIA RTX PRO 3000 Blackwell Generation Laptop GPU (12\,GB VRAM, SM120, compute capability 12.0, driver 591.59) (\texttt{docs/BENCHMARKS.md}).

\textbf{Build configuration.} Rust crates are compiled in \texttt{--release} mode. CUDA kernel modules are staged cubin-first with portable PTX fallback and explicit JIT warmup. Python components use \texttt{pyxlog} with CUDA-enabled PyTorch.

\textbf{Statistical protocol.} The benchmark harness uses Criterion.rs with the settings in \Cref{tab:stat-protocol} (\texttt{docs/BENCHMARKS.md}).

\begin{table}[htbp]
  \centering
  \caption{Criterion.rs benchmark settings.}
  \label{tab:stat-protocol}
  \small
  \begin{tabular}{@{}ll@{}}
    \toprule
    \textbf{Setting} & \textbf{Value} \\
    \midrule
    Sample size & 10--100 runs (benchmark-dependent) \\
    Warm-up & 3 iterations (PTX JIT, memory pool init) \\
    Significance level & 0.10 \\
    Noise threshold & 0.05 (ignore ${<}5\%$ variance) \\
    Seeding & Deterministic LCG (no system entropy) \\
    \bottomrule
  \end{tabular}
\end{table}

Artifact-backed values are quoted directly from checked-in JSON/Markdown evidence. Tables sourced from \texttt{docs/BENCHMARKS.md} are benchmark targets for the development machine rather than archived run logs. GPU measurements include explicit warm-up to amortize kernel-module loading and CUDA context establishment.

\subsection{Absolute Performance}

\subsubsection{Deterministic Subsystem}

For transitive closure, hash join, and aggregation in isolation the repository publishes target throughput envelopes (\Cref{app:contract}) rather than archived result tables, so we make no measured absolute-throughput claim for these operations here. The measured deterministic result reported in this paper is the worst-case-optimal join ablation below.

\subsubsection{Worst-Case Optimal Joins}

The WCOJ subsystem (\Cref{sec:datalog}) is measured against the binary-join baseline on four paper-scale fixtures (call-graph edges, Andersen points-to~\cite{andersen1994}, \texttt{ddisasm}-style disassembly~\cite{ddisasm}, and a neural-symbolic mining analog), each with 4{,}194{,}304 input rows. The first three are standard Datalog program-analysis shapes; the fourth is a recursive neural-symbolic rule-mining workload. We measure at this scale deliberately: WCOJ speedups depend strongly on input size and skew, and synthetic micro-benchmarks systematically overstate them, so multi-rule recursive workloads at millions of tuples are the honest setting for a reproducible claim. \Cref{tab:wcoj} reports the artifact-backed speedups; both paths are required to produce identical row sets before timing, and every measurement is a median over 10 runs with coefficient of variation ${\le}0.05$.

\begin{table}[htbp]
  \centering
  \caption{WCOJ vs.\ binary-join speedup on paper-scale fixtures (4.19M input rows each). Source: checked-in WCOJ ablation evidence.}
  \label{tab:wcoj}
  \small
  \begin{tabular}{@{}lr@{}}
    \toprule
    \textbf{Fixture} & \textbf{WCOJ speedup} \\
    \midrule
    \texttt{call\_graph\_edge} & $29.62\times$ \\
    \texttt{andersen}         & $26.96\times$ \\
    \texttt{ddisasm}          & $28.79\times$ \\
    \texttt{nesy\_analog}      & $26.60\times$ \\
    \midrule
    \textbf{Geometric mean}   & $\mathbf{27.96\times}$ \\
    \bottomrule
  \end{tabular}
\end{table}

These are single-system measurements of xlog's own WCOJ path against its own binary-join path on skewed inputs; on uniform or empty inputs the adaptive classifier falls back to the binary-join chain. They are an ablation, not a comparison against another engine---this paper makes no head-to-head claim against any external system (\Cref{sec:limits}). Peak observed allocation was roughly $2.29$\,GB, within the device's 12\,GB budget.

\subsubsection{Runtime Optimization Layer}

The runtime optimization layer (\Cref{sec:datalog}) is measured on dedicated repeated-session fixtures. The persistent hash-index manager records a $3.21\times$ speedup on a build-heavy repeated-session semi-join (cached median $0.079$\,s vs.\ uncached $0.255$\,s) against a $1.5\times$ target, with zero tracked device-to-host or host-to-device transfers in the checked-in persistent-index evidence. The profile-gated shared-memory chain scorer records a $5.58\times$ speedup on a chain-hot fixture, with a bounded ${-}1.3\%$ regression on the small-input control in the checked-in chain-scoring evidence. Common-subexpression elimination and adaptive re-optimization are validated for output equivalence and duplicate-work reduction with no added data-plane transfers, but are reported here as correctness-gated optimizations rather than headline speedups.

\subsubsection{Probabilistic Subsystem}

The probabilistic engine supports exact inference via Decision-DNNF knowledge compilation and GPU-accelerated Monte Carlo sampling. The repository publishes target latency and throughput envelopes for both (\Cref{app:contract}) but does not yet ship archived absolute-timing reports with stable per-phase numbers, so we make no measured absolute-latency claim for the engine in isolation. The measured probabilistic result in this paper is the circuit-caching ablation of the next subsection. The MC engine's hot-loop optimizations---targeted reset planning, bulk GPU sampling, and pointer-buffer reuse---are present in the implementation but unquantified here.

\subsubsection{Neural-Symbolic Subsystem}

Neural-symbolic training is measured on the \texttt{01\_minimal} benchmark (MNIST single-digit addition, 512 training images, 5 epochs, batch size 64). The timing summary below uses the checked-in neural comparison artifact; the cache-ablation study remains a separate 3-epoch measurement of the compilation-cache effect.

\begin{table}[htbp]
  \centering
  \caption{Training performance summary.}
  \label{tab:training-summary}
  \small
  \begin{tabularx}{\linewidth}{@{}l l X@{}}
    \toprule
    \textbf{Metric} & \textbf{Value} & \textbf{Source} \\
    \midrule
    PTX JIT cold start & $0.02$\,s & Cubin-first module loading target from \texttt{docs/BENCHMARKS.md} \\
    First epoch (cold) & $71.654$\,s & checked-in neural comparison, mean over 3 seeds \\
    Steady-state epoch (warm) & $0.248$\,s & checked-in neural comparison, mean epochs 2--5 \\
    Total training (5 epochs) & $72.645$\,s & checked-in neural comparison, mean over 3 seeds \\
    Per-query cost & $0.97$\,ms & checked-in neural comparison, mean forward+backward query cost \\
    Circuit cache speedup & $2.74\times$ & Cache ablation artifact, 95\% CI: $[2.29, 3.18]$ \\
    \bottomrule
  \end{tabularx}
\end{table}

The cache ablation study compared circuit-cached vs.\ uncached training over 3 seeds and 3 epochs each. Mean training time dropped from $242.9$\,s (uncached) to $88.9$\,s (cached), yielding the $2.74\times$ speedup reported in \Cref{sec:prob} (\texttt{examples/neural/results/evidence/cache\_ablation\_20260218.json}).

\textbf{DeepProbLog baseline.} The checked-in DeepProbLog comparison artifact is partial and protocol-mismatched, so it should be read as a qualitative baseline note rather than a clean head-to-head measurement. The strict full-train DeepProbLog configuration (2 epochs, batch size 2048) reached held-out addition accuracy of 23.24\% after 2:31:07, while a smaller limited run was manually aborted after 60 minutes. The same artifact reports XLOG training loss rather than held-out accuracy and marks one XLOG seed incomplete, so direct quantitative comparison remains premature.

\textbf{GA-based ILP.} The genetic algorithm reliability gate (\texttt{test\_ga\_reliability\_50}) achieved a $3.25\times$ wall-clock reduction from 1447\,s to 445.30\,s by reducing redundant attempts, with identical statistical quality: 200/200 success rate and Clopper--Pearson lower-95 CI $= 0.981725$ (\texttt{docs/evidence/2026-03-05-ga-runtime-closure/sweep-results.md}).

\subsection{Qualitative Comparison}

\Cref{tab:comparison} positions xlog relative to existing systems across key capability dimensions. Each system addresses a subset of the design space; xlog's contribution is unifying these capabilities within a single GPU-resident runtime.

\begin{table*}[htbp]
  \centering
  \caption{Capability comparison across probabilistic and GPU logic systems.}
  \label{tab:comparison}
  \small
  \begin{tabularx}{\linewidth}{@{}l X X X X@{}}
    \toprule
    \textbf{Dimension} & \textbf{DeepProbLog} & \textbf{ProbLog2} & \textbf{GPUlog} & \textbf{xlog} \\
    \midrule
    Symbolic execution & CPU (SWI-Prolog) & CPU (tabling) & GPU (HISA) & GPU (semi-naive fixpoint) \\
    Probabilistic inference & Exact (PySDD, CPU) & Exact + approx (CPU) & None & Exact (d-DNNF, GPU) + MC (GPU) \\
    Knowledge compilation & PySDD (CPU) & d-DNNF via external Decision-DNNF compilers (CPU) & None & external Decision-DNNF compile + GPU circuit eval \\
    Neural integration & PyTorch + Prolog bridge & None & None & Zero-copy DLPack, fused backward \\
    Zero-copy ML interop & No (CPU--GPU transfers) & No & No & Yes (DLPack, device-resident tensors) \\
    Differentiable ILP & Via TP-operator (CPU) & No & No & GPU-resident credit/loss path with evolving API \\
    \bottomrule
  \end{tabularx}
\end{table*}

\textbf{Published GPU Datalog results.} GPUlog~\cite{gpulog} reports up to $45\times$ speedup over CPU baselines using HISA (Hash-Indexed Sorted Array) indexing with lock-free deduplication. VFLog~\cite{vflog} demonstrates up to $200\times$ gains over CPU column engines with a column-oriented GPU Datalog runtime. Both systems are restricted to deterministic evaluation; neither supports probabilistic inference, knowledge compilation, or neural network integration. xlog draws on the same insight---that Datalog's relational algebra maps naturally to GPU parallelism---but extends the execution model through the probabilistic and neural-symbolic layers described in \Cref{sec:prob,sec:nesy}.

\textbf{Honest framing.} The comparison above is qualitative. xlog has not been benchmarked head-to-head against GPUlog or VFLog on identical workloads. The deterministic throughput targets (\Cref{app:contract}) represent xlog's own benchmark contract, not comparative claims, and the WCOJ and circuit-caching numbers are ablations against xlog's own baselines. The DeepProbLog comparison is limited to the \texttt{01\_minimal} task with differing dataset sizes, metrics, and partially completed archived runs, as noted above.

\subsection{CUDA Certification}

The CUDA kernel certification suite (\texttt{crates/xlog-cuda-tests/}) provides systematic coverage of all GPU kernel operations. In the May 2026 checked-in state, the statistics in \Cref{tab:cuda-cert} hold.

\begin{table}[htbp]
  \centering
  \caption{CUDA certification suite statistics (May 2026 checked-in state).}
  \label{tab:cuda-cert}
  \small
  \begin{tabular}{@{}ll@{}}
    \toprule
    \textbf{Metric} & \textbf{Value} \\
    \midrule
    Total tests             & 207 \\
    Categories              & 33 (25 CUDA infrastructure + 8 GPU-specific) \\
    Recorded full certification & 100\% (207/207) \\
    Kernel modules covered  & 22 \\
    \bottomrule
  \end{tabular}
\end{table}

The 25 CUDA infrastructure categories cover toolchain validation, launch configuration, pointer bounds, memory spaces (global, shared, local), synchronization, warp-level operations, control flow divergence, atomics, floating-point semantics, integer operations, determinism, async pipelines, caching, host-device transfer, multi-stream execution, multi-GPU, hardware stress, algorithmic correctness, blind spots, edge matrices, and float filter semantics (\texttt{docs/certification/2026-01-22-neural-symbolic-certification-report.md}). The 8 GPU-specific categories validate circuit forward evaluation, circuit backward (gradient) kernels, weight injection, transfer efficiency, circuit caching, PTX robustness, SAT/CDCL solving, and device-count invariants (\texttt{docs/architecture/cuda-certification.md}). The 22 covered runtime kernel modules correspond to the staged runtime CUDA kernel set plus the build-time ILP-family modules documented in the certification architecture notes. Because the public repository does not rely on hosted GPUs, the full certification suite is exercised through GPU-backed local release validation (\texttt{scripts/validate\_release\_gpu.sh --mode release}) and the manual self-hosted \texttt{cuda-ci} workflow; the always-on push/PR CI remains non-GPU and covers formatting, shell/workflow linting, package metadata, and no-GPU build checks.
