\section{Related Work}\label{sec:related}

xlog draws on and extends five lines of research: the tradition of logic programming languages, GPU-accelerated Datalog evaluation, probabilistic logic programming, GPU SAT solving, and differentiable inductive logic programming. We survey each in turn, positioning xlog's contributions relative to the state of the art.

\subsection{Logic Programming Languages}

xlog sits in a long line of logic programming languages but is the first to target GPU execution of the full language surface. Prolog~\cite{swiprolog} is the canonical universal logic language, offering unification and extensive metaprogramming, running on a CPU tree-walking interpreter. Mercury~\cite{mercury} introduced strong typing and mode analysis to logic programming, compiling to optimized native code; xlog inherits the typed-predicate idea but targets device kernels rather than CPU code. Souffl{\'e}~\cite{souffle} is a typed Datalog dialect that compiles to efficient C++ and is widely used for program analysis; it is CPU-resident. Answer-set programming with clingo~\cite{clingo} provides nonmonotonic reasoning via answer-set semantics on CPU grounders and solvers; xlog is not an ASP system, but its stratified semantics and probabilistic extensions intersect ASP applications. Probabilistic logic languages such as ProbLog~\cite{problog2} and DeepProbLog~\cite{deepproblog} express weighted facts and neural predicates as first-class syntactic constructs, but their reasoners run on the CPU while neural components run on the GPU. xlog unifies all of these---typed logic, Datalog evaluation, probabilistic facts, neural predicates---as one language, compiled and executed entirely on the GPU.

\subsection{GPU Datalog}

Prior GPU-native Datalog systems---gpulog~\cite{gpulog} and vflog~\cite{vflog}---target the Datalog evaluation backend that xlog also provides (\Cref{sec:datalog}), but do not surface a language of xlog's breadth (typed predicates, user-defined functions, modules, probabilistic and neural extensions). GPUlog ported bottom-up fixed-point computation to CUDA and reported approximately $45\times$ speedups over equivalent CPU Datalog engines on transitive-closure benchmarks. VFLog advances this line with a vertically-fused, column-oriented kernel design that avoids materializing intermediate relations, achieving up to $200\times$ speedups over CPU column engines on recursive graph queries. More recently, mnmgDatalog~\cite{mnmgdatalog} extends GPU Datalog to multi-node, multi-GPU clusters, distributing semi-naive evaluation across devices via partitioned fact stores and all-to-all delta exchanges. All three systems target deterministic Datalog: they implement relational algebra operators (join, project, deduplicate) on the GPU and accelerate fixed-point iteration, but none supports probabilistic semantics, weighted model counting, or gradient computation over derived facts. xlog shares VFLog's commitment to columnar storage and kernel fusion for relational operators, but extends the GPU-resident execution model to probabilistic and differentiable reasoning, maintaining a single address space from semi-naive evaluation through circuit-based inference to neural-symbolic gradient propagation. Where these prior GPU engines rely on binary joins, xlog additionally promotes eligible cyclic and multiway rules to a worst-case-optimal join path~\cite{ngo2018wcoj,veldhuizen2014leapfrog,wang2023freejoin} (\Cref{sec:datalog,sec:eval}), avoiding the intermediate-relation blowup that binary-join plans incur on skewed cyclic queries.

\subsection{Probabilistic Logic Programming}

ProbLog2~\cite{problog2} established the modern pipeline for exact probabilistic inference in logic programs: ground the relevant program, encode provenance as a propositional formula, compile to a tractable target (d-DNNF or SDD), and evaluate weighted model counts over the compiled circuit. DeepProbLog~\cite{deepproblog} extends ProbLog2 by replacing selected probabilistic facts with the outputs of neural networks, enabling end-to-end gradient-based training of neural predicates within a logical framework. NeurASP~\cite{neurasp} takes a related approach, integrating neural network outputs as probability annotations on answer-set programs and using stable-model semantics for inference. These systems demonstrate that combining neural perception with symbolic reasoning yields strong performance on tasks requiring both pattern recognition and logical structure. However, all three perform symbolic inference on the CPU: ProbLog2 invokes an external Decision-DNNF compiler as a host-side subprocess, DeepProbLog inherits this pipeline and adds a Python-level training loop that shuttles gradients between CPU logic and GPU tensors, and NeurASP relies on CPU-bound answer-set solving. The checked-in DeepProbLog baseline artifact in this repository is qualitative rather than a clean head-to-head benchmark: it records a 2:31:07 full-train run at 23.24\% held-out addition accuracy and a smaller aborted run, while also noting protocol mismatches versus XLOG's archived measurements. xlog eliminates the host/device bottleneck by compiling d-DNNF circuits into GPU-resident XGCF format with compile-once/evaluate-many semantics, executing both forward weighted model counting and backward gradient propagation as level-parallel CUDA kernels without host transfers during training iterations.

\subsection{GPU SAT}

ParaFROST~\cite{parafrost} is a GPU-accelerated CDCL SAT solver that parallelizes clause database simplification---bounded variable elimination, subsumption, and self-subsuming resolution---on the GPU while retaining sequential conflict-driven search on the CPU. It is a complete solver with proof-generation capabilities, competitive with MiniSat and CaDiCaL on industrial benchmarks. FastFourierSAT~\cite{fastfouriersat} takes a different approach, reformulating SAT as continuous optimization over Fourier coefficients and applying local search on the GPU; it is an incomplete solver suited to satisfiable instances but unable to certify unsatisfiability. Both systems are standalone solvers designed for the SAT competition setting, not embedded within a logic programming pipeline. xlog's GPU conflict-driven clause-learning (CDCL) module (\Cref{sec:prob}) shares ParaFROST's strategy of offloading clause manipulation to device kernels, but integrates the solver as a verification component within the knowledge compilation pipeline---checking circuit-formula equivalence via two UNSAT proofs, computing minimal unsatisfiable cores, and providing proof certificates that ensure the correctness of compiled circuits.

\subsection{Differentiable ILP}

Differentiable inductive logic programming (dILP)~\cite{dilp} demonstrates that first-order rule induction can be cast as a differentiable optimization problem: candidate rules are scored by soft forward-chaining over all possible rule instantiations, and gradient descent adjusts rule weights to maximize coverage of positive examples. The approach is elegant but computationally demanding: forward-chaining materializes a dense tensor over the full rule space at each step, scaling cubically with the number of constants in the domain. Subsequent work has reduced this cost through pruning and approximate materialization, but implementations remain CPU-based and struggle beyond a few hundred constants. xlog's dILP subsystem (\Cref{sec:nesy}) replaces dense rule materialization with sparse GPU mask computation: candidate rules are represented as sparse bitmasks over the ground-atom space, credit assignment is performed entirely on-device via scatter-gather operations, and a six-gate promotion pipeline controls rule acceptance. This sparse, GPU-resident formulation avoids the cubic blowup of dense materialization while preserving differentiability, enabling rule induction over domains with thousands of constants.

\subsection{Positioning}

Each system surveyed above excels in its niche. GPUlog and VFLog demonstrate that relational fixed-point computation maps naturally to GPU architectures. ProbLog2 and DeepProbLog prove that knowledge compilation provides a principled bridge between logic and learning. ParaFROST shows that clause-level parallelism yields practical SAT speedups. dILP establishes that rule induction is amenable to gradient-based optimization. xlog unifies these capabilities on a single GPU-resident platform with zero-copy ML interop---not claiming superiority on any single axis, but providing the integrated stack. By keeping fact stores, compiled circuits, solver state, and gradient buffers in device memory throughout the reasoning pipeline, xlog eliminates the host-device transfer overhead that fragments existing multi-system approaches, offering practitioners a single engine where deterministic evaluation, probabilistic inference, SAT verification, and differentiable learning compose without leaving the GPU.
