\section{Limitations and Future Work}\label{sec:limits}

\subsection{Current Limitations}

\textbf{NVIDIA GPU required.} xlog targets CUDA exclusively. The entire execution stack---from hash-join kernels and radix sort to Monte Carlo sampling and knowledge compilation---is written in CUDA C. There is no OpenCL, Metal, or ROCm backend. This is a direct consequence of the GPU-native design philosophy: rather than abstracting over heterogeneous devices behind a portability layer, xlog commits to a single GPU programming model to exploit CUDA-specific features such as cooperative groups, warp-level primitives, and unified virtual addressing. These primitives are load-bearing, not incidental: the persistent-index manager, the recorded-launch discipline, and the in-kernel hash-consing of the knowledge-compilation pipeline all depend on them, and a lowest-common-denominator portability layer---forcing extra synchronization points and disallowing warp operations---would erode the zero-transfer performance model that is the system's reason for existing. Portability is thus traded away deliberately. Users without an NVIDIA GPU cannot run xlog.

\textbf{All data must fit in GPU memory.} Relations, indices, circuit buffers, and Monte Carlo sample arrays are allocated entirely on-device. There is no out-of-core spilling mechanism; a program whose working set exceeds GPU VRAM will fail with a \texttt{RESOURCE\_EXHAUSTED} error rather than degrade silently. For most Datalog workloads this is not a bottleneck---a 24\,GB GPU holds billions of 8-byte tuples---but large-scale knowledge graphs or high-sample-count MC inference can hit the ceiling.

\textbf{Python batch-query helpers are typed but not float-complete.} The Python helper APIs---\texttt{batch\_fact\_membership}, \texttt{batch\_fact\_membership\_device}, \texttt{batch\_tagged\_credit}, and \texttt{batch\_tagged\_credit\_device}---now perform schema-aware uploads for \texttt{u32}, \texttt{u64}, \texttt{i32}, \texttt{i64}, and \texttt{bool} columns, so wide integer IDs no longer truncate silently. However, these helper paths still reject floating-point batch uploads (\texttt{f32}/\texttt{f64}) at the binding layer. The underlying engine and DLPack relation-import paths remain typed; this is a convenience-API gap rather than a core execution limit.

\textbf{Differentiable ILP is beta.} The dILP subsystem---structure learning via gradient-based rule search---shipped in v0.4.0 and received significant reliability and performance work from v0.5.0 through v0.9.0-rc. It remains beta: the search space is restricted to definite programs, convergence is sensitive to learning-rate schedules, and the API surface may change in minor releases.

\textbf{Epistemic and solver maturity (release candidate).} The epistemic runtime (\Cref{sec:epistemic}) executes accepted programs GPU-natively, but the v0.9.0 release candidate is not yet a stable release: the MaxSAT/portfolio solver services and the end-to-end probabilistic--epistemic evidence path are only partially implemented, and nested modal operators, epistemic integrity constraints, and rules coupling multiple epistemic predicates across components are rejected with typed diagnostics rather than supported. The accepted surface is deliberately minimal (\Cref{sec:epistemic}).

\textbf{No formal head-to-head benchmarks.} xlog's probabilistic inference pipeline occupies similar territory to DeepProbLog, but no controlled comparison on identical programs and datasets exists. The baseline experiments in \Cref{sec:eval} use different dataset sizes and evaluation protocols. Claims of speedup relative to CPU-based probabilistic logic systems remain informal until such benchmarks are published.

\subsection{Future Work}

The following directions remain outside the current released contract.

\textbf{Out-of-core execution.} For programs whose materialized relations exceed GPU memory, a streaming execution mode would partition relations across host and device, paging tiles through GPU memory in fixpoint-iteration order. This would remove the single-GPU-memory ceiling described above, at the cost of added host-device transfer overhead. The v0.8.6 persistent hash-index manager reduces repeated-session build cost but does not spill to host; out-of-core execution is targeted for v0.10.0 alongside multi-GPU partitioning (\texttt{ROADMAP.md}).

\textbf{Epistemic residency expansion.} The v0.9.x line implements an accepted EIR/GPU epistemic execution surface with FAEEL default semantics, G91 compatibility mode, finite nested modal-chain normalization, split/joint planning, tuple-key membership, and GPU-backed runtime paths. The remaining future work is stronger residency and coverage: device-resident/no-host-interaction WFS, broader non-finite tuple-key forms, and semantic forms outside the finite accepted release surface.

\textbf{Multi-GPU partitioned evaluation.} Inspired by mnmgDatalog's radix-hash partitioning and GPU-aware all-to-all shuffle, a multi-GPU backend would distribute relations across devices and coordinate join and union operations via NVLink or PCIe. Partitioning kernels are described in the roadmap but unimplemented; this is targeted for v0.10.0 (\texttt{ROADMAP.md}).
